---
title: "Lab 09 Halloween Candy"
author: "Omar Siddiqui"
output: 
  pdf_document:
    latex_engine: xelatex
---
## Importing candy data

```{r}
# We read in the CSV and set row.names=1 so that the candy names become
# row labels instead of their own column. This lets us reference candies
# by name later (e.g. candy["Twix",])
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

## Q1. How many different candy types are in this dataset?
```{r}
# nrow() counts the number of rows, and since each row is one candy type,
# this gives us the total number of candies in the dataset
nrow(candy)
```

## Q2. How many fruity candy types are in the dataset?
```{r}
# The fruity column contains 1s and 0s (1 = fruity, 0 = not fruity)
# sum() adds up all the 1s, which gives us the count of fruity candies
sum(candy$fruity)
```

## What is your favorite candy?
```{r}
# We can index into the dataframe using the candy name as a row label
# and then select the winpercent column with $winpercent
# winpercent = the % of head-to-head matchups this candy won
candy["Starburst", ]$winpercent
```

## Q3. My favorite candy is Starburst and its winpercent value is:
```{r}
# Starburst is a classic fruity candy - lets see how it stacks up!
candy["Starburst", ]$winpercent
# A winpercent around 67 means Starburst won about 67% of its matchups -
# pretty popular!
```

## Q4. Winpercent for Kit Kat
```{r}
candy["Kit Kat", ]$winpercent
# Kit Kat scores very high - the crispy wafer + chocolate combo is a winner
```

## Q5. Winpercent for Tootsie Roll Snack Bars
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
# Tootsie Rolls score much lower - not many people's favorite
```

## Q6-Q7: Skim the data
```{r}
# skimr gives us a fast summary of every column - great for getting a feel
# for the data before diving into analysis. Think of it like a supercharged summary()
library("skimr")
skim(candy)
```
# Q6: winpercent is on a completely different scale than all other columns.
# Everything else ranges from 0 to 1, but winpercent goes from 0 to 100.
# This matters a lot for PCA later - we'll need to scale the data!

# Q7: For candy$chocolate, 0 means the candy does NOT contain chocolate
# and 1 means it DOES contain chocolate. These are called "binary" or
# "dummy" variables - a common way to encode yes/no information numerically.

## Q8. Histogram of winpercent
```{r}
# A histogram is always a good first step to understand how a variable
# is distributed - is it spread out? Skewed? Are there outliers?

# Base R version - quick and simple
hist(candy$winpercent, 
     main = "Distribution of Candy Win Percentages",
     xlab = "Win Percent",
     col = "lightblue")

# ggplot2 version - more customizable
library(ggplot2)
ggplot(candy) +
  aes(x = winpercent) +
  geom_histogram(bins = 30, fill = "lightblue", color = "white") +
  labs(title = "Distribution of Candy Win Percentages",
       x = "Win Percent",
       y = "Count")
```

# Q9: The distribution is NOT symmetrical - it is slightly right skewed,
# meaning there are more candies with lower win percentages and a few
# very popular candies pulling the tail to the right

# Q10: The center of the distribution is BELOW 50%, meaning most candies
# lose more matchups than they win - only a few candies are truly beloved

## Q11-Q12. Chocolate vs Fruity - which is more popular?
```{r}
# as.logical() converts the 0s and 1s to FALSE/TRUE so we can use them
# as a filter to subset the winpercent values for each candy type

# Average winpercent for chocolate candies
mean(candy$winpercent[as.logical(candy$chocolate)])

# Average winpercent for fruity candies
mean(candy$winpercent[as.logical(candy$fruity)])

# t.test checks whether the difference between these two averages is
# statistically significant or could just be due to random chance
# If p-value < 0.05, the difference is considered statistically significant
t.test(candy$winpercent[as.logical(candy$chocolate)],
       candy$winpercent[as.logical(candy$fruity)])
```
# Q11: Chocolate candy is ranked HIGHER on average than fruity candy
# Chocolate avg ~60% vs fruity avg ~44% - a pretty big gap!

# Q12: YES this difference IS statistically significant
# The p-value is much less than 0.05, meaning this difference is very
# unlikely to be due to random chance. Chocolate really does win more.

## Q13. Five least liked candies
```{r}
# arrange() sorts the dataframe, head() takes the first n rows
# Since we want LEAST liked, we sort in ascending order (default)
library(dplyr)
candy %>%
  arrange(winpercent) %>%
  head(5)
# Note: these are the candies that lost the most matchups
```

## Q14. Top 5 favorite candies
```{r}
# desc() reverses the sort order so highest winpercent comes first
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
# Note: Reese's dominates the top spots - peanut butter + chocolate is king
```

## Q15. First barplot - rough draft
```{r}
# This first plot will be ugly - that's okay! We build plots iteratively,
# starting simple and improving step by step. This is normal in data science.
library(ggplot2)
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
# Problem: bars are in alphabetical order which makes it hard to compare
```

## Q16. Barplot sorted by winpercent
```{r}
# reorder() sorts the y-axis (candy names) by winpercent values
# Now we can actually see which candies are most/least popular at a glance
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

## Adding color to show candy type
```{r}
# We create a color vector with one color per candy
# Start with all black, then overwrite with colors for specific types
# This is a clever way to add a categorical color coding without ggplot legends
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"  # brown-ish
my_cols[as.logical(candy$bar)] = "brown"             # darker brown
my_cols[as.logical(candy$fruity)] = "pink"           # pink for fruity

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols)

# Now we can visually see that chocolate candies (brown) cluster at the top
# and fruity candies (pink) tend to be lower ranked
```
# Q17: The worst ranked chocolate candy is Sixlets
# Q18: The best ranked fruity candy is Starburst - go figure, our favorite!

## Q19-Q20. Win percent vs Price percent scatter plot
```{r}
# ggrepel automatically moves text labels so they don't overlap each other
# This is super useful when you have many labeled points close together
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5) +
  labs(title = "Candy Win % vs Price %",
       x = "Win Percent (popularity)",
       y = "Price Percent (cost)")
# We want candies in the TOP LEFT: high win%, low price% = best value
```
# Q19: Reese's Miniatures offer the most bang for your buck -
# very popular but relatively inexpensive

# Q20: The top 5 most expensive candies are Nik L Nip, Nestle Smarties,
# Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate.
# Nik L Nip is the least popular of these despite being the most expensive!

## Q21. Lollipop chart of pricepercent
```{r}
# A lollipop chart is a cleaner alternative to a bar chart
# geom_segment draws the stick, geom_point draws the dot at the end
# Much less visual clutter than filled bars when comparing many items
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent),
                   xend = 0), col = "gray40") +
  geom_point() +
  labs(title = "Candy Price Rankings",
       x = "Price Percent",
       y = "")
# Notice how many candies share the same price point - the dots stack up
```

## Q22-Q23. Correlation matrix
```{r}
# A correlation matrix shows how every variable relates to every other variable
# Values close to +1 = strong positive correlation
# Values close to -1 = strong negative correlation (anti-correlated)
# Values close to 0 = little to no relationship
# corrplot visualizes this as a colored grid - much easier to read than numbers
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
# Q22: Chocolate and fruity are the most ANTI-correlated (dark red/orange)
# This makes intuitive sense - candies are rarely both chocolatey AND fruity

# Q23: Chocolate and winpercent (and bar and winpercent) are most positively
# correlated - confirming what we saw in our earlier t-test and bar charts

## Q24. Principal Component Analysis (PCA)
```{r}
# PCA reduces many variables down to a few "principal components" that
# capture the most important patterns of variation in the data
# scale=TRUE is CRITICAL here because winpercent is on a 0-100 scale
# while everything else is 0-1. Without scaling, winpercent would dominate
# the analysis just because its numbers are bigger, not because it's more important
pca <- prcomp(candy, scale = TRUE)
summary(pca)
# PC1 explains ~36% of variance, PC2 explains ~11% - together they capture
# nearly half of all the variation in the dataset in just 2 dimensions!

# Basic PCA plot
plot(pca$x[,1:2], col = my_cols, pch = 16)
# We can already see chocolate candies (brown) clustering on the left
# and fruity candies (pink) clustering on the right

# Make a nicer ggplot version
# First we combine our PCA results with the original candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
  aes(x = PC1, y = PC2,
      size = winpercent/100,  # bigger points = more popular candy
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p

# Add labels with ggrepel so we can identify each candy
library(ggrepel)
p + geom_text_repel(size = 3.3, col = my_cols, max.overlaps = 7) +
  theme(legend.position = "none") +
  labs(title = "Halloween Candy PCA Space",
       subtitle = "Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (pink), other (black)",
       caption = "Data from 538")
# Candies that appear close together in this plot are "similar" to each other
# based on all their combined properties

# Loadings plot - shows which original variables drive each PC
# This tells us WHAT makes candies different from each other along each axis
ggplot(pca$rotation) +
  aes(PC1, PC2) +
  geom_label(label = rownames(pca$rotation)) +
  labs(title = "PCA Loadings",
       x = "PC1 loadings",
       y = "PC2 loadings")
```
# Q24: Fruity, hard, and pluribus are picked up strongly by PC1 in the
# POSITIVE direction. This makes sense - fruity candies tend to be hard
# and come in bags with multiple pieces (pluribus). Chocolate, bar, and
# winpercent load strongly in the NEGATIVE PC1 direction, meaning chocolate
# bar candies are on the opposite end of the spectrum from fruity hard candies.
# We saw this exact relationship highlighted in the correlation matrix earlier!

## Q25. Summary - what makes a winning candy?
# Based on all three analyses:
# VISUALIZATION: The colored barplot clearly showed chocolate candies
# dominating the top of the rankings
#
# CORRELATION: The correlation matrix confirmed chocolate and winpercent
# are strongly positively correlated, while fruity and winpercent are not
#
# PCA: The PCA score plot showed that the most popular candies (largest points)
# cluster on the negative PC1 side, which the loadings plot tells us corresponds
# to chocolate, bar, and caramel characteristics
#
# CONCLUSION: A winning candy is most likely to be a chocolate candy bar
# with additional components like caramel, peanuts/almonds, or crisped rice.
# Think Reese's, Twix, Kit Kat, Snickers - the classics dominate for a reason!

## Q26. Are popular candies more expensive?
```{r}
losers <- candy[which(candy$winpercent < 50),]
winners <- candy[which(candy$winpercent >= 50),]

mean(winners$pricepercent)
mean(losers$pricepercent)

# t.test to check if the price difference is statistically significant
t.test(winners$pricepercent, losers$pricepercent)
```
# Winners have a higher average price (0.580) vs losers (0.374)
# The p-value is 0.0006, which is well below 0.05, so this difference
# IS statistically significant. Popular candies DO tend to cost more -
# if you want the good stuff, you have to pay for it!


## Q27. Are candies with more sugar more popular?
```{r}
mean(winners$sugarpercent)
mean(losers$sugarpercent)

t.test(winners$sugarpercent, losers$sugarpercent)
```
## Winners have slightly higher sugar content (0.534) vs losers (0.431)
# But the p-value is 0.094, above 0.05, so NOT statistically significant
# Interesting! Sugar content alone does NOT drive candy popularity.
# It's really about the chocolate, not the sugar. Quality over quantity!