---
title: "Lab 11: Protein Structure Prediction"
author: "Omar Siddiqui"
format: gfm
execute:
  echo: true
---

## Setup
```{r}
library(bio3d)
library(pheatmap)
library(jsonlite)

setwd("/Users/omar/BIMM-143/Lab 11")
```


## Loading AlphaFold Results

```{r}
results_dir <- "/Users/omar/BIMM143/Lab 11/hivprdimer_23119/"

pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

basename(pdb_files)
```

## Aligning and Superposing Structures

```{r}
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
pdbs
```

## RMSD Analysis

We can use RMSD to measure structural distance between all pairs of models.

```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

```{r}
colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

## Plotting pLDDT Scores

The pLDDT score is a per-residue confidence metric ranging from 0-100. Values above 70 are considered high confidence.

```{r}
pdb <- read.pdb("1hsg")

plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

## Core Structure Analysis

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

## RMSF Analysis

RMSF measures conformational variance along the structure.

```{r}
rf <- rmsf(xyz)
plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Aligned Error (PAE)

PAE values help us assess confidence in the relative positions of different domains.

```{r}
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)

pae1 <- read_json(pae_files[1], simplifyVector = TRUE)
pae5 <- read_json(pae_files[5], simplifyVector = TRUE)

attributes(pae1)
head(pae1$plddt)
```

```{r}
# Compare max PAE scores - lower is better
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae1$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
# Re-plot model 1 with same scale for fair comparison
plot.dmat(pae1$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

## Residue Conservation Analysis

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                       full.names = TRUE)

aln <- read.fasta(aln_file[1], to.upper = TRUE)
dim(aln$ali)
```

```{r}
sim <- conserv(aln)

plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

## Mapping Conservation to Structure

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```