---
title: "Lab 5: Data Exploration and Visualization in R"
author: "Mohammad Omar Siddiqui"
format: pdf
---

## Setup and Package Installation

Before plotting, we need to ensure our packages are installed and loaded. 

```{r setup, message=FALSE}
# install.packages("ggplot2")
# install.packages("gapminder")
# install.packages("dplyr")

library(ggplot2)
```

## Basic Scatter Plots with built-in data

First, we will visualize the built-in `cars` dataset. 

```{r cars-plot}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point()
```

Now, we add a trendline to see the relationship more clearly.

```{r cars-multi-geom}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Reading External Data and Customizing Plots

Next, we will load a gene expression dataset to see how drugs affect specific cell lines.

```{r read-data}
url <- "[https://bioboot.github.io/bimm143_S26/class-material/up_down_expression.txt](https://bioboot.github.io/bimm143_S26/class-material/up_down_expression.txt)"
genes <- read.delim(url)
head(genes)

Now we plot this data and customize the colors based on whether the genes are upregulated, downregulated, or unchanging.

```{r custom-plot}
ggplot(genes, aes(x = Condition1, y = Condition2, col = State)) +
  geom_point() +
  scale_color_manual(values = c("blue", "gray", "red")) +
  labs(title = "Gene Expression Changes Upon Drug Treatment",
       subtitle = "Scatter plot of Condition 1 vs Condition 2",
       x = "Control (No Drug)",
       y = "Drug Treatment",
       color = "Expression State") +
  theme_minimal()
```

## Exploring Gapminder and Flipping Bar Charts

Finally, we bring in the dataset charting global population, wealth, and life expectancy.

```{r gapminder-data}
library(gapminder)
library(dplyr)

gapminder_americas <- gapminder %>%
  filter(year == 2007, continent == "Americas")
```

```{r bar-plot}
ggplot(gapminder_americas, aes(x = reorder(country, lifeExp), y = lifeExp)) +
  geom_col(fill = "steelblue") +
  coord_flip() + 
  labs(title = "Life Expectancy in the Americas (2007)",
       x = "Country",
       y = "Life Expectancy (Years)") +
  theme_minimal()
```