---
title: "Lab 12 Section 4: Population Scale Analysis"
author: "Omar Siddiqui"
date: 11 May 2026
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q13

```{r}
# Read the data
expr <- read.table("rs8067378_ENSG00000172057.6.txt", header = TRUE)

# Sample size for each genotype
table(expr$geno)

# Median expression for each genotype
tapply(expr$exp, expr$geno, median)
```

The sample sizes are: A/A (n=108), A/G (n=233), and G/G (n=121). 
The corresponding median expression levels are: A/A = 31.25, A/G = 25.07, 
and G/G = 20.07. There is a clear decrease in median expression with each 
additional copy of the G allele.

## Q14

```{r}
# Boxplot of expression by genotype
boxplot(exp ~ geno, data = expr,
        xlab = "Genotype (rs8067378)",
        ylab = "ORMDL3 Expression",
        main = "Expression by Genotype for SNP rs8067378",
        col = c("tomato", "lightgreen", "steelblue"))
```

From the boxplot, A/A individuals have the highest ORMDL3 expression, A/G is 
intermediate, and G/G has the lowest. This indicates that the SNP rs8067378 
does affect ORMDL3 expression — each copy of the G allele is associated with 
roughly 5-6 units lower expression. The asthma-risk G allele therefore appears 
to reduce ORMDL3 transcription in an additive manner.
